Estimation of Manning's Friction Coefficient using DART Buoys Data during \tohoku Tsunami

### 2014-01-02 paper/coefs.tex
Cleaned up tsunami section and added some references

\subsection{Inferring Manning's Friction Coefficient}
 \label{sec:manning}

We seek to infer the Manning's $n$ coefficients from water surface elevation
data measured at different gauges as shown in Figure~\ref{fig:setup}.
Bayesian inference can be applied directly to infer the uncertain parameters
using Equation~\eqref{eq:post}. In our case, the observation data $\vec d$ 
are the water surface elevation at the gauges;
their model predicted counterparts $\vec G$ is predicted by \geoclaw.
The uncertain parameters $\vec \theta$ will be denoted by $\{n_i\}_{i=1}^3$ for the
three Manning's coefficients treated as the unknown parameters. 
In this case  Equation~\eqref{eq:post} can be written as:

>>>>

\subsubsection{Inferring Manning's Friction Coefficient}
\label{sec:infmanning}

We seek to infer the Manning's $n$ coefficients from water surface elevation
data measured at different gauges as shown in Figure~\ref{fig:setup_buoy_data}.
Bayesian inference can be applied directly to infer the uncertain parameters
using Equation~\eqref{eq:post}. In our case, the observation data $\vec d$ 
are the water surface elevation at the gauges;
their model predicted counterparts $\vec G$ is predicted by \geoclaw.
The uncertain parameters $\vec \theta$ will be denoted by $\{n_i\}_{i=1}^3$ for the
three Manning's coefficients treated as the unknown parameters. 
In this case  Equation~\eqref{eq:post} can be written as:

### 2014-01-02 paper/pcstudy.tex
Cleaned up tsunami section and added some references

Figure~\ref{fig:rlzs} shows the evolution of the
water surface elevation predicted at the four different gauges 
using \geoclaw for the 125 different realizations 
required to compute the PC expansions. We notice that the 
variability in water surface elevation is insignificant for the first 
hour at all gauges as the plots of the different realizations superimpose.
Later in time, the variability starts to increase with time at the different gauges 
as indicated from the thickness of the bands formed by the plots of the realizations.
This variability appears to be significant at $t\sim$ 5400~s, 6000~s, 3600~s and 7200~s
for gauges 21401, 21413, 21418, 21419 respectively.
This is consistent with the distance from the gauge to the source of the earthquake
(epicenter located approximately 72 km east of \tohoku) where gauge 21418 is the closest and gauge 21419 is the farthest as shown in Figure~\ref{fig:setup}.
The uncertainty in the prediction of water surface  elevation continues till the end of the simulations
at all gauges.

>>>>

Figure~\ref{fig:rlzs} shows the evolution of the
water surface elevation predicted at the four different gauges 
using \geoclaw for the 125 different realizations 
required to compute the PC expansions. We notice that the 
variability in water surface elevation is insignificant for the first 
hour at all gauges as the plots of the different realizations superimpose.
Later in time, the variability starts to increase with time at the different gauges 
as indicated from the thickness of the bands formed by the plots of the realizations.
This variability appears to be significant at $t\sim$ 5400~s, 6000~s, 3600~s and 7200~s
for gauges 21401, 21413, 21418, 21419 respectively.
This is consistent with the distance from the gauge to the source of the earthquake
(epicenter located approximately 72 km east of \tohoku) where gauge 21418 is the closest and gauge 21419 is the farthest as shown in Figure~\ref{fig:setup_buoy_locations}.
The uncertainty in the prediction of water surface  elevation continues till the end of the simulations
at all gauges.

### 2014-01-02 paper/tsunami.tex
Cleaned up tsunami section and added some references

\section{Simulation Setup}

In this section we briefly describe the forward computational model used, \geoclaw, and the case study used in the following uncertainty analysis, the great \tohoku earthquake with observations from NOAA DART buoys.

\subsection{\geoclaw} \label{ssub:geoclaw}

The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:
\begin{equation} \label{eq:swe}
    \begin{aligned}
    &\pd{}{t} h + \pd{}{x} (hu) + \pd{}{y} (hv) = 0, \\
    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| hu \\
    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| hv
    \end{aligned}
\end{equation}
where $h$ is the depth of the water column, $u$ and $v$ the velocities in the 
longitudinal and latitudinal directions respectively, $g$ the acceleration due 
to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom 
friction coefficient.  As is common in tsunami modeling, $C_f$ is calculated 
given a Manning's $n$ parameterization such that:
\begin{equation}
    C_f = \frac{g n^2}{h^{5/3}}
\end{equation}
where $n$ is an empirically determined parameter known as Manning's $n$.

@@

\subsection{The \tohoku Tsunami} \label{ssub:tohoku}

The \tohoku earthquake and tsunami of 2011 has been the subject of a number of studies due to the wealth of observational evidence and severity of the tsunami.  The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and caused massive devastation across Japan.  In the case study presented, the simulation is based on work done in \cite{MacInnes:2013cr} and modified to use the specified variable friction field.

\subsubsection{Source Model}

The MacInnes \emph{et al.} study found that source models from tsunami
inversions were

Since the previous study \cite{MacInnes:2013cr} did not find that the source model effected the tsunami wave-forms at the DART buoys, we did not expect there to be much difference between the available models.  After comparing the same analysis with two different source models, namely the models from \cite{Ammon:2011dm} and .

[4] Ammon used a seismic and GPS inversion technique
[7] Saito et al. used a tsunami inversion

Since the previous study found that the source model used did not impact the off-shore gauge stations, the ones used here from \cite{Ammon:2011dm}, was chose the source model that was commonly available.  Access to the finest bathymetry located near the Japanese coastline was also not included as it primarily effected only the shore observations.

\subsubsection{Bathymetry}
The bathymetry used in the simulation is a combination of ETOPO 1' and 4' 
accurate bathymetry for the region \cite{Amante:2009ud}.  The additional 
bathymetry data used in the MacInnes \emph{et al.} was useful only near the
shore and had a smaller effect on the regions near the observations used in
this study.

\subsubsection{Observations}
The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).  The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.  The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).

\section{Manning’s $n$ Friction Parameterization}
%We use the widely used .  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.

Manning's $n$ law provides a relationship between the friction coefficient appearing in the shallow
water equations governing tsunamis and the roughness parameter $n$. 
This Manning's $n$ coefficient is variable that is highly spatial 
and dependent on the surface characteristics of the sea-floor. 
The direct measurement of $n$ is therefore impossible, instead, values
are often assigned from tables that are empirically estimated
for common types of land that are determined from satellite images.
As a result, the Manning’s n coefficients used in tsunami models
often contain large amounts of uncertainty.

The uncertainty in the Manning's $n$ coefficient leads to uncertainties
in the predicted quantities of interest such as the water elevations
and becomes a health risk during tsunami events. In addition, 
the prediction of these quantities of interest are often highly sensitive to 
the changes in these uncertain parameters. 
In this work, we seek to quantify parametric uncertainty namely due 
to the uncertainty in the roughness parameter $n$. To this end,
we characterize the roughness $n$ by three different parameters
that are varied in three different regions but are assumed
constant in each region. The three regions are as follows: 
on-shore whose coefficient is denoted by $n_1$, near-shore 
denoted by $n_2$, and deep-water denoted by $n_3$.
The three regions are shown in Figure~\ref{fig:ceofs}.
To carry out the uncertainty quantification we have employed polynomial chaos expansions 
as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
to our problem is described in Section~\ref{sec:uqpce}.  
To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $n \in [0.005-0.2]$.

Note that on the Sendai plain a Manning's $n$ value of 0.035 was used to account for pasture, farmland and rice paddies.

>>>>

\section{The \tohoku Tsunami} 
\label{sec:tohoku}

The \tohoku earthquake and tsunami of 2011 has been the subject of a number of
studies due to the wealth of observational evidence and severity of the tsunami.
The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and
caused massive devastation across Japan.  In the case study presented, the
simulation is based on work done in \cite{MacInnes:2013cr} and modified to use
the specified variable friction field. In this section we briefly describe the
forward computational model used, \geoclaw, the  simulation setup used in the
uncertainty analysis and finally describe the observations from NOAA DART buoys.

\subsection{\geoclaw} \label{ssub:geoclaw}
The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:
\begin{equation} \label{eq:swe}
    \begin{aligned}
    &\pd{}{t} h + \pd{}{x} (hu) + \pd{}{y} (hv) = 0, \\
    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| hu \\
    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| hv
    \end{aligned}
\end{equation}
where $h$ is the depth of the water column, $u$ and $v$ the velocities in the 
longitudinal and latitudinal directions respectively, $g$ the acceleration due 
to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom 
friction coefficient.  As is common in tsunami modeling, $C_f$ is calculated 
given a Manning's $n$ parameterization such that:
\begin{equation}
    C_f = \frac{g n^2}{h^{5/3}}
\label{eq:coef}
\end{equation}
where $n$ is an empirically determined parameter known as Manning's $n$.

@@

\subsection{DART Buoys Observations}

The observations used in this study are from the Deep-ocean Assessment and
Reporting of Tsunamis (DART) buoy system developed and maintained by the
National Oceanic and Atmospheric Administration (NOAA).  The purpose of the
network is to provide early-warning detection and forecasting of tsunami
propagation in the Pacific Ocean \cite{Milburn:1996wm}.  The DART buoys closest
to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418,
and 21419 whose locations are shown in Figure~\ref{fig:setup_buoy_locations} and
whose de-tided data for the event are shown in Figure~\ref{fig:setup_buoy_data}.
The time $t=0$ is set to be the start of the earthquake.  We can see that the
primary tsunami waves generated reach the gauges between 40 minutes and 2 hours
after the earthquake.  The large spikes later in time present in the data at
gauge 21418 are data anomalies and not subsequent tsunami waves.

\subsection{Bathymetry and Earthquake Source Model}

The bathymetry used in the simulation is a combination of ETOPO 1' and 4'
accurate bathymetry for the region \cite{Amante:2009ud}.  The additional finer
resolution bathymetry data used in MacInnes \emph{et al.} were not used in this
study since the simulations were not sensitive to the finer bathymetry at the
DART buoy locations in the deep-water.

Two earthquake source models were run to test whether the friction UQ analysis
was sensitive to different rupture models, the first is from Saito et al.
\cite{Saito:2011bh} and is based on an inversion analysis based on dispersive
tsunami simulations.  The second is from Ammon \emph{et al.} \cite{Ammon:2011dm} and
used inversion of the seismic waves to reconstruct the rupture.  These two
models were chosen as they take very different approaches to the reconstruction
of the original fault and were shown to obtain the best un-shifted match to the
observations at the DART buoys for their respective types of reconstruction.  As
discussed in MacInnes \emph{et al.}, the source models used do not include rupture
timing and this could have an impact on the accuracy of the simulation.

\comment{Still working on getting the Saito \emph{et al.} source model, seems to do better at the DART buoys.}

\section{Manning’s $n$ Friction Parameterization} \label{sec:manning}

Manning's $n$ law provides a relationship between the friction coefficient
\eqref{eq:coef} appearing in the shallow water equations \eqref{eq:swe}
governing tsunamis and the roughness parameter $n$.  This Manning's $n$
coefficient is variable that is highly spatial  and dependent on the surface
characteristics of the sea-floor.  The direct measurement of $n$ is therefore
impossible, instead, values are often assigned from tables that are empirically
estimated for common types of land that are determined from satellite images. As
a result, the Manning’s n coefficients used in tsunami models often contain
large amounts of uncertainty.

The uncertainty in the Manning's $n$ coefficient leads to uncertainties in the
predicted quantities of interest such as the water elevations and becomes a
health risk during tsunami events. In addition,  the prediction of these
quantities of interest are often highly sensitive to  the changes in these
uncertain parameters.  In MacInnes \emph{et al.} a Manning's $n$ coefficient of 0.025
was used except on the Sendai plane where 0.035 was used to account for pasture,
farmland and rice paddies.  In this work, we seek to quantify parametric
uncertainty namely due to the uncertainty in the roughness parameter $n$. To
this end, we characterize the roughness $n$ by three different parameters that
are varied in three different regions but are assumed constant in each region.
The three regions are as follows:  on-shore whose coefficient is denoted by
$n_1$, near- shore  denoted by $n_2$, and deep-water denoted by $n_3$. The three
regions are shown in Figure~\ref{fig:ceofs}. To carry out the uncertainty
quantification we have employed polynomial chaos expansions as used in
\cite{sraj:2013a,sraj:2013b}.  The method and its application to our problem is
described in Section~\ref{sec:uqpce}.  To represent the uncertainty in the
Manning's $n$ coefficients a uniform distribution was assumed as the prior where
$N$ was sampled in the interval $n \in [0.005-0.2]$.
